---
title: "data_tidying"
author: "Ben Gerber"
date: "11/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
```

# Read and clean data

```{r}
catch_original <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", stringsAsFactors = FALSE)
head(catch_original)
```

Remove `all` column and `notes` column because we don't need them.

```{r}
catch_data <- catch_original %>% 
  select(-c(All, notesRegCode))
```

```{r}
summary(catch_data)
#table(catch_data$Chinook)
```

## Fix Chinook Column

Use `mutate` to fix Chinook Column

```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>%
  mutate(Chinook  = as.numeric(Chinook))
```

Find rows that got turned into NA

```{r}
i <- which(is.na(catch_clean$Chinook))
i
```

```{r}
catch_original[i, ]
```

# Reshape our data

```{r}
catch_long <- catch_clean %>%
  pivot_longer(cols = -c(Region, Year), names_to = "Species", values_to = "Count")
head(catch_long)
```

```{r}
catch_wide <- catch_long %>%
  pivot_wider(names_from = Year, values_from = Count)
head(catch_wide)
```

`rename` count column to `catch_thousands`

```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = Count)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
  mutate(catch = catch_thousands * 1000) %>%
  select(-catch_thousands)

head(catch_long)
```

# Summarize our data

```{r}
mean_region <- catch_long %>%
  group_by(Region, Species) %>%
  summarize(catch_mean = mean(catch),
            num_obs = n())

mean_region
```

Calculate total catch per species

```{r}
total_catch <- catch_long %>%
  group_by(Species) %>%
  summarise(sum = sum(catch)) %>%
  arrange(desc(sum))

total_catch
```

```{r}
total_species_pre_1900 <- catch_long %>%
  filter(Year < 1900) %>%
  group_by(Species) %>%
  summarise(total = sum(catch)) %>%
  arrange(desc(total))

total_species_pre_1900
```

# Join to Region Table

Read in data table

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1",stringsAsFactors = FALSE) %>% 
    select(code, mgmtArea) %>%
    rename (Region = code)

head(region_defs)
```

```{r}
#catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))
catch_joined <- left_join(catch_long, region_defs, by = "Region")
head(catch_joined)
```

# Separate and Unite

```{r}
site_codes <- data.frame(site = c("HAB-100",
                                  "HAW-101",
                                  "OAH-102",
                                  "OAH-103",
                                  "MAI-100"),
                         stringsAsFactors = FALSE)
```

Separate island and site number using `separate`

```{r}
site_codes_split <- site_codes %>%
  separate(site, into = c("island", "site_number"), sep = '-', remove = FALSE)

head(site_codes_split)
```

Unite back again

```{r}
united <- site_codes_split %>%
  unite(site_code, island, site_number, sep = '_')

head(united)
```


### End